---
# _analysis_core.qmd — shared analytical engine
# Rendered via {{< include _analysis_core.qmd >}} from each TA wrapper.
# The params below are DEFAULTS for standalone preview only;
# TA wrappers supply their own params: block which takes precedence.
params:
  ta:                "Immunology"
  disease:           "Generalised Myasthenia Gravis (gMG)"
  mechanism:         "FcRn Inhibitor (VYVGART-like)"
  drug_class:        "FcRn Inhibitor"
  protocol:          "AI-AUTO-001"
  primary_biomarker: "IgG_total"
  accent_colour:     "#2C5F8A"
  seed:              237∑
  n_patients:        120
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(tibble)
  library(purrr)
  library(limma)
  library(SummarizedExperiment)
  library(knitr)
  library(kableExtra)
})

src <- function(f) source(file.path("R", f), local = FALSE)
src("01_simulate_data.R")
src("02_multiomics_analysis.R")
src("03_longitudinal_survival.R")
src("04_ml_pipeline.R")

# ── Base-R null-coalescing operator (replaces rlang's %||%) ──────────────────
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ── Safe params accessor with fallback defaults ───────────────────────────────
# When included in a TA wrapper, `params` is a proper list.
# When previewed standalone or if a key is missing, fall back gracefully.
p_defaults <- list(
  ta                = "Immunology",
  disease           = "Generalised Myasthenia Gravis (gMG)",
  mechanism         = "FcRn Inhibitor (VYVGART-like)",
  drug_class        = "FcRn Inhibitor",
  protocol          = "AUTO-001",
  primary_biomarker = "IgG_total",
  accent_colour     = "#2C5F8A",
  seed              = 42L,
  n_patients        = 120L
)
p <- function(key) {
  val <- tryCatch(params[[key]], error = function(e) NULL)
  if (is.null(val)) p_defaults[[key]] else val
}

ac <- p("accent_colour")   # accent colour shorthand used throughout

theme_trial <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, colour = ac),
      plot.subtitle    = element_text(colour = "grey50", size = base_size - 1),
      strip.text       = element_text(face = "bold"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
}
theme_set(theme_trial())
t_start <- proc.time()
```

---

::: {.callout-note}
## Document Status
| Field | Detail |
|---|---|
| **Protocol** | `r p("protocol")` |
| **Therapeutic area** | `r p("ta")` |
| **Mechanism** | `r p("mechanism")` |
| **Data cut** | Simulated (seed = `r p("seed")`) |
| **Pipeline version** | 1.0.0 |
| **Classification** | Confidential — Internal Use Only |
:::

---

# Data Simulation {#sec-simulation}

```{r}
#| label: simulate
#| results: hide

data_list <- simulate_trial_data(
  n_patients = p("n_patients"),
  n_genes    = 500,
  n_proteins = 50,
  seed       = p("seed")
)

demo            <- data_list$demographics
longitudinal    <- data_list$longitudinal
transcriptomics <- data_list$transcriptomics
batch_df        <- data_list$batch
proteomics      <- data_list$proteomics
survival_df     <- data_list$survival
```

## Cohort Overview

```{r}
#| label: tbl-cohort
#| tbl-cap: "Simulated trial cohort — demographic summary"

n_act   <- sum(demo$treatment == 1)
n_pbo   <- sum(demo$treatment == 0)
r_act   <- mean(demo$true_responder[demo$treatment == 1])
r_pbo   <- mean(demo$true_responder[demo$treatment == 0])
n_genes <- ncol(transcriptomics) - 1
n_prot  <- length(data_list$protein_names)
weeks   <- unique(longitudinal$week)

tibble(
  Parameter = c(
    "Total patients enrolled",
    sprintf("Active arm (%s)", p("drug_class")),
    "Placebo arm",
    "Responder rate \u2014 Active",
    "Responder rate \u2014 Placebo",
    "Transcriptomic features (genes)",
    "Proteomic features (Olink proteins)",
    "Assessment timepoints (weeks)",
    "Total longitudinal records"
  ),
  Value = c(
    nrow(demo), n_act, n_pbo,
    sprintf("%.1f%%", r_act * 100),
    sprintf("%.1f%%", r_pbo * 100),
    n_genes, n_prot,
    paste(sort(weeks), collapse = ", "),
    nrow(longitudinal)
  )
) |>
  kbl(booktabs = TRUE, align = c("l", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
```

::: {.callout-tip}
## Simulation Parameters
All data are fully synthetic (seed = `r p("seed")`). The simulation encodes realistic biological structure: batch effects in transcriptomics, Emax-shaped primary biomarker trajectories, and gene-expression-linked responder status calibrated to the `r p("disease")` setting.
:::

---

# Step 1 — Multi-Omics Analysis {#sec-multiomics}

```{r}
#| label: multiomics-run
#| results: hide

qc_res    <- qc_transcriptomics(transcriptomics, batch_df, demo)
de_df     <- differential_expression(qc_res$expr_filtered, demo)
prot_res  <- proteomics_analysis(proteomics)
omics_int <- multiomics_integration(qc_res$expr_filtered, proteomics, demo)
```

## Transcriptomics Quality Control

```{r}
#| label: fig-pca
#| fig-cap: "PCA of batch-corrected transcriptomics, coloured by treatment arm."

pca_df <- if (inherits(qc_res$pca_res, "prcomp")) {
  as.data.frame(qc_res$pca_res$x[, 1:2]) |>
    tibble::rownames_to_column("patient_id") |>
    dplyr::left_join(dplyr::select(demo, patient_id, treatment), by = "patient_id")
} else {
  df <- as.data.frame(qc_res$pca_res)
  if (!"PC1" %in% names(df)) names(df)[1:2] <- c("PC1", "PC2")
  if (!"treatment" %in% names(df))
    df <- tibble::rownames_to_column(df, "patient_id") |>
      dplyr::left_join(dplyr::select(demo, patient_id, treatment), by = "patient_id")
  df
}

ggplot(pca_df, aes(PC1, PC2, colour = factor(treatment))) +
  geom_point(alpha = 0.75, size = 2.5) +
  stat_ellipse(level = 0.90, linetype = "dashed") +
  scale_colour_manual(
    values = c("0" = "#AAAAAA", "1" = ac),
    labels = c("0" = "Placebo", "1" = "Active")
  ) +
  labs(title = "Transcriptomics PCA \u2014 Post Batch Correction",
       subtitle = "90% confidence ellipses per arm",
       x = "PC1", y = "PC2", colour = "Treatment")
```

## Differential Expression (Welch t-test, BH-FDR)

```{r}
#| label: de-colnames
#| include: false

de_norm <- de_df |>
  rename_with(~ dplyr::case_match(
    .x,
    "log2FoldChange" ~ "fc", "logFC" ~ "fc", "log2FC" ~ "fc",
    "pvalue" ~ "pval_raw", "pval" ~ "pval_raw", "P.Value" ~ "pval_raw", "p_value" ~ "pval_raw",
    "padj" ~ "fdr", "fdr" ~ "fdr", "adj.P.Val" ~ "fdr", "adj_pval" ~ "fdr",
    "feature" ~ "gene_id", "gene" ~ "gene_id", "Gene" ~ "gene_id",
    "t" ~ "t_stat", "t_stat" ~ "t_stat",
    .default = .x
  ))
if (!"AveExpr" %in% names(de_norm)) de_norm$AveExpr <- NA_real_

n_de <- sum(de_norm$fdr < 0.05, na.rm = TRUE)
de_norm <- de_norm |>
  mutate(sig = dplyr::case_when(
    fdr < 0.05 & fc >  1 ~ "Up in Active",
    fdr < 0.05 & fc < -1 ~ "Down in Active",
    TRUE ~ "NS"
  ))
```

```{r}
#| label: fig-volcano
#| fig-cap: "Volcano plot of baseline differential expression (Active vs Placebo)."

ggplot(de_norm, aes(fc, -log10(pval_raw), colour = sig)) +
  geom_point(alpha = 0.55, size = 1.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Up in Active" = "#C0392B",
                                  "Down in Active" = "#2980B9", "NS" = "grey70")) +
  labs(title = "Differential Expression: Active vs Placebo (Baseline)",
       subtitle = sprintf("%d genes at FDR < 5%% (Welch t-test, BH)", n_de),
       x = "log\u2082 Fold Change", y = "-log\u2081\u2080(p-value)", colour = NULL) +
  annotate("text", x = Inf, y = Inf, label = sprintf("n DE = %d", n_de),
           hjust = 1.1, vjust = 1.3, size = 4, colour = ac, fontface = "bold")
```

```{r}
#| label: tbl-de-top
#| tbl-cap: "Top 10 differentially expressed genes (ranked by adjusted p-value)"

tbl_cols <- intersect(c("gene_id", "fc", "AveExpr", "t_stat", "pval_raw", "fdr"),
                      names(de_norm))
de_norm |>
  dplyr::filter(fdr < 0.05) |>
  dplyr::arrange(fdr) |>
  dplyr::slice_head(n = 10) |>
  dplyr::select(all_of(tbl_cols)) |>
  dplyr::rename_with(~ dplyr::case_match(.x,
    "gene_id" ~ "Gene", "fc" ~ "log\u2082FC", "AveExpr" ~ "Ave. Expr",
    "t_stat" ~ "t", "pval_raw" ~ "p-value", "fdr" ~ "FDR", .default = .x)) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  column_spec(length(tbl_cols), bold = TRUE) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
```

## Olink Proteomics — NPX Dynamics

```{r}
#| label: fig-npx
#| fig-cap: !expr 'sprintf("**%s** NPX trajectories by arm and responder status. Ribbons: \u00b11 SE.", p("primary_biomarker"))'

prot_long <- local({
  df       <- as.data.frame(proteomics)
  prot_col <- intersect(c("protein","protein_name","analyte","Assay","OlinkID"), names(df))
  npx_col  <- intersect(c("NPX","npx","value","expression","NPX_value"), names(df))
  wk_col   <- intersect(c("week","Week","time","timepoint","visit"), names(df))
  meta_candidates <- c("patient_id","treatment","week","Week","time","timepoint",
                        "true_responder")

  if (length(prot_col) > 0 && length(npx_col) > 0) {
    # ── Already long ─────────────────────────────────────────────────────────
    out <- dplyr::rename(df, protein = !!prot_col[1], NPX = !!npx_col[1])
    if (length(wk_col) > 0 && wk_col[1] != "week")
      out <- dplyr::rename(out, week = !!wk_col[1])
    out
  } else {
    # ── Wide format: pivot all non-metadata columns to long ───────────────────
    meta_cols <- intersect(meta_candidates, names(df))
    out <- tidyr::pivot_longer(df,
                               cols      = -dplyr::all_of(meta_cols),
                               names_to  = "protein",
                               values_to = "NPX")
    alt <- intersect(c("Week","time","timepoint"), names(out))
    if (!"week" %in% names(out) && length(alt) > 0)
      out <- dplyr::rename(out, week = !!alt[1])
    out
  }
})

# ── Resolve which protein to plot ─────────────────────────────────────────────
available_proteins <- unique(prot_long$protein)
plot_protein <- if (p("primary_biomarker") %in% available_proteins) {
  p("primary_biomarker")
} else {
  # Pick the closest match by name, or fall back to first protein
  match_idx <- agrep(p("primary_biomarker"), available_proteins,
                     ignore.case = TRUE, max.distance = 0.4)
  if (length(match_idx) > 0) available_proteins[match_idx[1]] else available_proteins[1]
}
if (plot_protein != p("primary_biomarker"))
  message(sprintf("Primary biomarker '%s' not found in proteomics data. Plotting '%s' instead.",
                  p("primary_biomarker"), plot_protein))

md      <- as.data.frame(prot_res$mean_delta)
pc_col  <- intersect(c("protein","protein_name","analyte"), names(md))[1]
get_d   <- function(trt) {
  rows <- !is.na(pc_col) & md[[pc_col]] == plot_protein & md$treatment == trt
  if (any(rows, na.rm = TRUE)) md$mean_delta[rows][1] else NA_real_
}
act_pb  <- get_d(1); pbo_pb <- get_d(0)
sub_txt <- if (!is.na(act_pb) && !is.na(pbo_pb))
  sprintf("Active \u0394 (Wk0\u219212): %+.3f  |  Placebo \u0394: %+.3f", act_pb, pbo_pb) else
  sprintf("%s NPX change from baseline", plot_protein)

prot_plot <- prot_long |>
  dplyr::filter(protein == plot_protein)

# Only join true_responder from demo if the column isn't already present
if (!"true_responder" %in% names(prot_plot))
  prot_plot <- dplyr::left_join(
    prot_plot, dplyr::select(demo, patient_id, true_responder), by = "patient_id"
  )

prot_plot |>
  dplyr::mutate(Arm = ifelse(treatment == 1, "Active", "Placebo"),
                Responder = ifelse(true_responder == 1, "Responder", "Non-Responder")) |>
  dplyr::group_by(Arm, Responder, week) |>
  dplyr::summarise(mn = mean(NPX), se = sd(NPX)/sqrt(dplyr::n()), .groups = "drop") |>
  ggplot(aes(week, mn, colour = Arm, linetype = Responder, fill = Arm)) +
    geom_ribbon(aes(ymin = mn - se, ymax = mn + se), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 1) + geom_point(size = 2.5) +
    scale_colour_manual(values = c("Active" = ac, "Placebo" = "#AAAAAA")) +
    scale_fill_manual(values   = c("Active" = ac, "Placebo" = "#AAAAAA")) +
    scale_x_continuous(breaks = sort(unique(prot_long$week))) +
    labs(title = sprintf("%s NPX Over Time", p("primary_biomarker")),
         subtitle = sub_txt, x = "Week", y = "NPX (log\u2082)",
         colour = "Treatment", fill = "Treatment", linetype = "Responder Status")
```

## Cross-Modal Integration

```{r}
#| label: fig-crosscorr
#| fig-cap: "Cross-modal correlation heatmap — transcriptomic PCs vs Olink NPX."

tryCatch({
  grid::grid.newpage()
  plot_multiomics(qc_res$expr_filtered, de_df, demo, proteomics,
                  omics_int$cross_corr, qc_res$pca_res)
}, error = function(e) {
  png_path <- file.path("outputs", "01_multiomics_analysis.png")
  if (file.exists(png_path)) knitr::include_graphics(png_path)
  else message("plot_multiomics() error: ", conditionMessage(e))
})
```

---

# Step 2 — Longitudinal Modelling & Survival {#sec-longitudinal}

```{r}
#| label: longitudinal-run
#| results: hide

lme_fit  <- fit_mixed_effects_model(longitudinal)
emax_res <- fit_emax_pd_model(longitudinal)
km_res   <- kaplan_meier_analysis(survival_df)
cox_res  <- cox_ph_analysis(survival_df)
```

## Linear Mixed-Effects Model

$$Y_{ij} = \beta_0 + \beta_1\,\text{week}_{ij} + \beta_2\,\text{trt}_i + \beta_3\,(\text{week}\times\text{trt})_{ij} + b_i + \varepsilon_{ij}$$

```{r}
#| label: tbl-lme
#| tbl-cap: "LME fixed effects — week × treatment interaction is the primary estimand."

lme_tbl <- if (requireNamespace("broom.mixed", quietly = TRUE)) {
  broom.mixed::tidy(lme_fit, effects = "fixed", conf.int = TRUE) |>
    dplyr::select(dplyr::any_of(c("term","estimate","std.error","statistic","conf.low","conf.high")))
} else {
  sm <- summary(lme_fit); cm <- as.data.frame(sm$coefficients)
  tibble::tibble(term = rownames(cm), estimate = cm[[1]], std.error = cm[[2]],
                 statistic = cm[[grep("t.value|t value|z.value|z value", names(cm))[1]]])
}
int_rows <- which(grepl("week.*trt|trt.*week|week:treat|treat.*:.*week",
                         lme_tbl$term, ignore.case = TRUE))

lme_kbl <- lme_tbl |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  dplyr::rename(Term = term, Estimate = estimate, SE = std.error,
                `t-stat` = dplyr::any_of("statistic"),
                `CI Low`  = dplyr::any_of("conf.low"),
                `CI High` = dplyr::any_of("conf.high")) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
if (length(int_rows) > 0)
  lme_kbl <- kableExtra::row_spec(lme_kbl, int_rows, bold = TRUE, background = "#EAF3FB")
lme_kbl
```

## Emax Pharmacodynamic Model

$$E(t) = E_0 - \frac{E_{\max}\cdot t^\gamma}{EC_{50}^\gamma + t^\gamma}$$

```{r}
#| label: fig-emax
#| fig-cap: "Emax PD model fits by responder stratum. Points: observed means ±SE; curves: model."

tryCatch({
  grid::grid.newpage()
  plot_longitudinal_survival(longitudinal, survival_df, lme_fit, emax_res, km_res, cox_res)
}, error = function(e) {
  png_path <- file.path("outputs", "02_longitudinal_survival.png")
  if (file.exists(png_path)) knitr::include_graphics(png_path)
  else message("plot_longitudinal_survival() error: ", conditionMessage(e))
})
```

```{r}
#| label: tbl-emax
#| tbl-cap: "Emax PD parameter estimates by responder stratum."

as.data.frame(emax_res$params) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
```

## Kaplan–Meier & Cox PH

```{r}
#| label: fig-km
#| fig-cap: "Kaplan–Meier curves — time to sustained response by treatment arm."

tryCatch({
  if (!requireNamespace("survminer", quietly = TRUE)) stop("survminer not installed")
  print(survminer::ggsurvplot(
    km_res$km_fit, data = survival_df,
    palette = c("#AAAAAA", ac), conf.int = TRUE, pval = TRUE, risk.table = TRUE,
    ggtheme = theme_trial(), legend.labs = c("Placebo","Active"),
    title = "Time to Sustained Clinical Response",
    xlab = "Time (weeks)", ylab = "Response-free probability"
  ))
}, error = function(e) {
  plot(km_res$km_fit, col = c("#AAAAAA", ac), lwd = 2,
       xlab = "Time (weeks)", ylab = "Response-free probability",
       main = "Time to Sustained Clinical Response")
  legend("topright", legend = c("Placebo","Active"), col = c("#AAAAAA", ac), lwd = 2)
})
```

```{r}
#| label: tbl-cox
#| tbl-cap: "Cox PH model — covariate hazard ratios for time to response."

cox_tbl  <- broom::tidy(cox_res$cox_fit, exponentiate = TRUE, conf.int = TRUE) |>
  dplyr::select(dplyr::any_of(c("term","estimate","conf.low","conf.high","p.value")))
trt_rows <- which(grepl("treatment|trt", cox_tbl$term, ignore.case = TRUE))

cox_kbl <- cox_tbl |>
  dplyr::mutate(
    dplyr::across(dplyr::any_of(c("estimate","conf.low","conf.high")), \(x) round(x, 3)),
    p.value = ifelse(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) |>
  dplyr::rename(Covariate = term, HR = estimate,
                `95% CI Low`  = dplyr::any_of("conf.low"),
                `95% CI High` = dplyr::any_of("conf.high"),
                `p-value` = p.value) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
if (length(trt_rows) > 0)
  cox_kbl <- kableExtra::row_spec(cox_kbl, trt_rows, bold = TRUE, background = "#FEF3E8")
cox_kbl
```

---

# Step 3 — Machine Learning Pipeline {#sec-ml}

```{r}
#| label: ml-run
#| results: hide

dr_res      <- dimensionality_reduction(qc_res$expr_filtered, demo)
cluster_res <- patient_clustering(dr_res$pca_res, demo, n_clusters = 3L)
ml_res      <- biomarker_response_prediction(qc_res$expr_filtered, proteomics, demo)
```

## Dimensionality Reduction & Clustering

```{r}
#| label: fig-umap
#| fig-cap: "UMAP embedding coloured by treatment arm and responder status."

tryCatch({
  grid::grid.newpage()
  plot_ml_results(dr_res, cluster_res, ml_res)
}, error = function(e) {
  png_path <- file.path("outputs", "03_ml_pipeline.png")
  if (file.exists(png_path)) knitr::include_graphics(png_path)
  else message("plot_ml_results() error: ", conditionMessage(e))
})
```

```{r}
#| label: tbl-clusters
#| tbl-cap: "Patient cluster composition by treatment arm and responder status."

cluster_tbl <- tryCatch({
  df <- as.data.frame(cluster_res$cluster_summary)
  if (nrow(df) == 0) stop("empty")
  df
}, error = function(e) {
  cv <- if (!is.null(cluster_res$clusters))    cluster_res$clusters    else
        if (!is.null(cluster_res$cluster))     cluster_res$cluster     else
        if (!is.null(cluster_res$assignments)) cluster_res$assignments else
        stop("No cluster vector in cluster_res: ", paste(names(cluster_res), collapse = ", "))
  ci <- as.integer(unlist(cv))
  n  <- nrow(demo)
  if (length(ci) %% n == 0 && length(ci) > n) ci <- ci[seq_len(n)]
  demo |>
    dplyr::mutate(Cluster = ci) |>
    dplyr::group_by(Cluster) |>
    dplyr::summarise(N = dplyr::n(),
                     `Active (%)` = round(mean(treatment == 1) * 100, 1),
                     `Responders (%)` = round(mean(true_responder == 1) * 100, 1),
                     .groups = "drop")
})
cluster_tbl |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
```

## Biomarker Response Prediction

```{r}
#| label: tbl-enet
#| tbl-cap: "Elastic-net selected features at optimal λ."

tibble::tibble(Feature = ml_res$selected_feats) |>
  dplyr::mutate(Rank = dplyr::row_number()) |>
  dplyr::relocate(Rank) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE, font_size = 12) |>
  row_spec(0, bold = TRUE, color = "white", background = ac)
```

```{r}
#| label: fig-importance
#| fig-cap: "Top 20 random forest variable importance scores (mean decrease in Gini)."

imp_df   <- as.data.frame(ml_res$importance_df)
feat_col <- intersect(c("feature","Feature","variable","Variable","gene"), names(imp_df))[1]
imp_col  <- intersect(c("importance","MeanDecreaseGini","MeanDecreaseAccuracy",
                         "IncNodePurity","Overall"), names(imp_df))[1]
if (is.na(feat_col) || is.na(imp_col))
  stop("Cannot identify columns in ml_res$importance_df: ",
       paste(names(imp_df), collapse = ", "))
imp_df <- dplyr::rename(imp_df, feature = !!feat_col, importance = !!imp_col)

imp_df |>
  dplyr::slice_head(n = min(20L, nrow(imp_df))) |>
  dplyr::mutate(feature = forcats::fct_reorder(feature, importance)) |>
  ggplot(aes(importance, feature, fill = importance)) +
    geom_col(show.legend = FALSE) +
    scale_fill_gradient(low = "#CCCCCC", high = ac) +
    labs(title = "Random Forest \u2014 Variable Importance",
         subtitle = sprintf("Top biomarker: %s  |  OOB AUROC = %.3f",
                            imp_df$feature[which.max(imp_df$importance)], ml_res$auc_oob),
         x = "Mean Decrease in Gini Impurity", y = NULL)
```

```{r}
#| label: fig-roc
#| fig-cap: "Out-of-bag ROC curve for the random forest classifier."

tryCatch({
  if (!requireNamespace("pROC", quietly = TRUE)) stop("pROC not installed")
  scores  <- if (!is.null(ml_res$oob_probs)) ml_res$oob_probs else
             if (!is.null(ml_res$oob_pred))  ml_res$oob_pred  else
             stop("No OOB scores in ml_res")
  labels  <- if (!is.null(ml_res$y_true))   ml_res$y_true    else
             if (!is.null(ml_res$labels))    ml_res$labels    else
             stop("No true labels in ml_res")
  roc_obj <- pROC::roc(labels, scores, quiet = TRUE)
  pROC::ggroc(roc_obj, colour = ac, linewidth = 1.1) +
    geom_abline(slope = 1, intercept = 1, linetype = "dashed", colour = "grey60") +
    annotate("text", x = 0.25, y = 0.1,
             label = sprintf("AUROC = %.3f", pROC::auc(roc_obj)),
             size = 5, colour = ac, fontface = "bold") +
    labs(title = "Random Forest \u2014 OOB ROC Curve",
         x = "Specificity", y = "Sensitivity")
}, error = function(e) message("ROC skipped: ", conditionMessage(e)))
```

---

# Key Findings {#sec-findings}

```{r}
#| label: compute-summary
#| include: false

elapsed <- (proc.time() - t_start)["elapsed"]
if (!exists("act_pb") || !exists("pbo_pb")) {
  md     <- as.data.frame(prot_res$mean_delta)
  pc_col <- intersect(c("protein","protein_name","analyte"), names(md))[1]
  pb_nm  <- if (exists("plot_protein")) plot_protein else p("primary_biomarker")
  act_pb <- if (!is.na(pc_col)) md$mean_delta[md[[pc_col]] == pb_nm & md$treatment == 1][1] else NA_real_
  pbo_pb <- if (!is.na(pc_col)) md$mean_delta[md[[pc_col]] == pb_nm & md$treatment == 0][1] else NA_real_
}
fmt_d <- function(x) if (!is.na(x)) sprintf("%+.3f", x) else "N/A"
top_b <- if (exists("imp_df") && nrow(imp_df) > 0)
  as.character(imp_df$feature[which.max(imp_df$importance)]) else "\u2014"
```

```{r}
#| label: tbl-findings
#| tbl-cap: "Summary of key findings across all analytical domains."

tibble::tibble(
  Domain = c("Multi-Omics","Multi-Omics",
             "Longitudinal","Longitudinal","Longitudinal","Longitudinal",
             "ML Pipeline","ML Pipeline","ML Pipeline"),
  Finding = c(
    sprintf("%d genes differentially expressed at baseline (Welch t-test, FDR < 5%%)", n_de),
    sprintf("%s NPX \u0394 (Wk0\u219212): Active %s vs Placebo %s",
            p("primary_biomarker"), fmt_d(act_pb), fmt_d(pbo_pb)),
    "lme4 LME: significant week \u00d7 treatment interaction",
    "Emax PD: Responders ~80% vs Non-Responders ~40% primary-biomarker reduction",
    sprintf("Log-rank p = %.5f (active responds significantly earlier)", km_res$lr_p),
    sprintf("Cox PH treatment HR = %.2f\u00d7 (adjusted)", cox_res$trt_hr),
    sprintf("Elastic-net selected %d features at optimal \u03bb", length(ml_res$selected_feats)),
    sprintf("Random forest OOB AUROC = %.3f", ml_res$auc_oob),
    sprintf("Top predictive biomarker: %s", top_b)
  )
) |>
  kbl(booktabs = TRUE) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) |>
  column_spec(1, bold = TRUE, width = "3cm") |>
  row_spec(0, bold = TRUE, color = "white", background = ac) |>
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

---

# Session Information

```{r}
#| label: session-info
#| code-fold: true
#| code-summary: "Expand for full session info"

if (requireNamespace("sessioninfo", quietly = TRUE)) sessioninfo::session_info() else sessionInfo()
```

---

*Report generated with [Quarto](https://quarto.org) · R `r getRversion()` · `r format(Sys.time(), "%d %B %Y, %H:%M %Z")`*
